Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-05-24 Paper/paper.tex
added HP's comments to intro

The Advanced LIGO (aLIGO) observatories completed their first observing run ``O1''
early-2016, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, they made the first terrestrial observation of gravitational 
waves~\cite{LIGOVirgo2016a}. Emitted by a pair of coalescing black holes, these
waves heralded an era of observational GW astrophysics as they traveled 
through Earth.
% 
By the end of the decade, we expect aLIGO to reach its design sensitivity,
as well as for Advanced Virgo and KAGRA to begin observing. 
% 
At its design sensitivity, we expect aLIGO to observe $\sim 70$ compact binary
mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf,LIGOVirgo2016a}. An important sub-class of these
are NSBH binaries, of which we expect $\mathcal{O}(10)$ GW observations
every year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.

The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about binary neutron stars (BNS)~\cite{
Lee1999a,Lee1999b,Lee2000,oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,
stergioulas:11,East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. While single 
observations are not as informative, Bayesian studies have shown that with
$20-30$ BNS observations, aLIGO would begin to distinguish between hard,
moderate and soft candidate equations of state~\cite{DelPozzo:13,
Chatziioannou:2015uea,Agathos:2015a}. These studies rely on the PN 
description of BNS dynamics, which is reasonable since BNSs merge at very high
frequencies and most of their signal power resides in inspiral.
% 
On the other hand, NSBHs merge at relatively lower frequencies than BNS, 
and their disruption during late-inspiral occurs at frequencies where aLIGO
has relatively more resolving power. This, combined with a shorter inspiral,
means that the merger portion of NSBH signals is relatively more valuable for
measuring $\lambdans$. Some past NSBH studies aimed at measuring tidal effects,
however, have used PN inspiral-only waveforms~\cite{Maselli:2013rza}. In doing
so, however, (i) they ignore the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms could easily dominate over the tidal terms 
themselves~\cite{Yagi:2014}, since there are unknown point-particle terms at 
orders {\it lower} than that at which tidal effects enter. Other studies that
use complete numerical simulations or derived models either use Fisher matrix 
to estimate $\lambdans$ measurement error~\cite{Lackey2011,Lackey:2013axa},
which is not reliable at realistic signal-to-noise ratios 
(SNR)~\cite{Vallisneri:2007ev}, or are limited in the binary parameter space
they sample~\cite{Foucart:2013psa}.

In this paper we study the measurability of neutron star's tidal deformability
from GWs emitted by {\it disruptive} NSBH mergers using aLIGO. We also probe how tidal 
effects affect the estimation of other binary parameters for the same class
of systems. This study improves upon previous work in many ways. First, we
include tidal effects during inspiral and merger in a consistent way, by using
the waveform model of Ref.~\cite{Lackey:2013axa} (henceforth ``LEA+''). Second,
we include the effect
of black hole spin on tidal GW signals, in addition to the effect of BH mass 
and NS's deformability itself. Third, we perform a fully Bayesian 
analysis, instead of using a Fisher matrix approximation. Fourth, we explore 
how our measurement errors decrease as we gain information from multiple 
(realistic) events.

We first probe the effect of not including tidal effects in parameter 
estimation templates on the recovery of non-tidal binary parameters, such as 
component masses and spins. This is the case of current and planned aLIGO
efforts.
To do so, we use the LEA+ model to generate a set of realistic signals, and
use that as data from which we estimate binary mass and spin parameters with
a Bayesian Monte-Carlo algorithm, using non-tidal (BBH) waveforms as filters.
% 
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
% 
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
%
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.

Next, we study the ability of aLIGO to constrain neutron star tidal 
deformability with a single observation of coalescing NSBHs. For this, we
use the same setup and parameter choices for signal waveforms as before, but
replace the template model with one that includes tidal effects from inspiral
through to merger~\cite{Lackey:2013axa} (i.e. LEA+). First of all, this reduces
the biases in the measurement of non-tidal parameters. Second, as we can see
from Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, at signal-to-noise
ratios $\simeq 20-30$, only under the most favorable circumstance (including a
BH with mass within the ``mass-gap'') can we put a factor of $1.5-2$ bound on
$\lambdans$ at $90\%$ confidence level. For similar signals with 
$\rho\gtrsim 30$ we can constrain $\lambdans$ within $\pm 50\%$ of its true
value. For more realistic SNRs and BH parameters, a single measurement is 
for $\lambdans$ is accompanied with a $\sim\pm150+\%$. uncertainty. While these
results are expected, multiple realistic observations is the next window that 
we explore in order to obtain more information on $\lambdans$. These results 
are detailed in Sec.~\ref{s1:PEwithNS}.

Multiple realistic observations deliver far more information than what a single
observation furnishes, as has been shown the case to be for binary neutron
stars~\cite{DelPozzo:13}. As our final study, we perform a fully-Bayesian 
analysis of how our estimation of $\lambdans$ changes as we accumulate more and
more detections. The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
%
\prayush{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
}
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}. In addition, we probe two paradigms, one
that allows for BH masses to lie within the astrophysical mass-gap (paradigm
A), and one that does not (paradigm B).
% 
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.

Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$. Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it provides only an 
amplitude model which needs to be (in future) augmented with a compatible
phase model. In addition, accuracy of the underlying numerical simulations 
used to calibrate the waveform model used here have not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and its effect
on our results. they are, therefore, limited by the limitations of our
waveform model. However, we expect the combined effect of modeling errors to
not change our qualitative conclusions.

>>>>

The Advanced LIGO (aLIGO) observatories completed their first observing run ``O1''
early-2016, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, they made the first terrestrial observation of gravitational 
waves~\cite{LIGOVirgo2016a}. Emitted by a pair of coalescing black holes, these
waves heralded an era of observational GW astrophysics as they traveled 
through Earth.
% 
Towards the end of this decade, we expect aLIGO to reach its design sensitivity.
In addition to the US-based efforts, we also expect the French-Italian detector Advanced
Virgo~\cite{aVIRGO,aVirgo2}, Japanese detctor KAGRA~\cite{kagra,Somiya:2011np},
and LIGO-India~\cite{2013IJMPD..2241010U} to begin observing at comparable
sensitivities within a few years. With a global network of sensitive GW
observatories, we can expect GW astronomy to face significant developments over the
coming years.
% 
% At design sensitivity, we expect aLIGO to observe $\sim 70$ compact binary
% mergers a year~\cite{Abadie:2010cf}.

\red{
Text about CBCs. How was each type observed so far?
}
\prayush{
Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf,LIGOVirgo2016a}.
% 
Compact binary merger rate estimates suggest
a GW detection rate of approximately a few tens of binary black holes
(BBH) every year~\cite{Abadie:2010cfa}. The actual masses of
astrophysical black holes are uncertain, but observations and
population synthesis studies suggest that BHs formed from stellar
core-collapse can have masses up to and higher than
$34M_\odot$~\cite{2008ApJ...678L..17S,2010ApJ...714.1217B}.  Also,
recent measurements using continuum fitting and X-ray reflection
fitting suggest that black holes can have high spin, with the BH
angular momentum in dimensionless units exceeding
$0.8$~\cite{2011ApJ...742...85G,2012MNRAS.424..217F,Gou:2013dna,
  2009ApJ...697..900M,McClintock:2006xd,Miller:2009cw,McClintock:2013vwa}.
Therefore the observations of GWs emitted by spinning BBHs will allow us to
understand the spin-spin and spin-orbit dynamics of the two-body
system, apart from allowing us to test strong-field dynamics of
General Relativity.
}
%

\red{
Text describing BHNS disruption and accretion dynamics.
}
An important sub-class of these
are NSBH binaries, of which we expect $\mathcal{O}(10)$ GW observations
every year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
%

\red{
Text focusing on tides
}
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.

The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible.
% 
\red{
Text about disruption of NSBHs, MOVE EARLIER.
}
\prayush{
The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.
}

\red{
Insert paragraph describing impact on GWs
}

A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about binary neutron stars (BNS)~\cite{
Lee1999a,Lee1999b,Lee2000,oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,
stergioulas:11,East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. 
Studies~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a} have shown that
with $20-30$ BNS observations, aLIGO would begin to distinguish between hard,
moderate and soft candidate equations of state. These studies rely on the PN 
description of BNS dynamics, which is reasonable since BNSs merge at very high
frequencies and most of their signal power resides in inspiral.
% 
On the other hand, NSBHs merge at relatively lower frequencies than BNS, 
and their disruption during late-inspiral occurs at frequencies where aLIGO
has relatively more resolving power. This, combined with a shorter inspiral,
means that the merger portion of NSBH signals is relatively more valuable for
measuring $\lambdans$. Some past NSBH studies aimed at measuring tidal effects,
however, have used PN inspiral-only waveforms~\cite{Maselli:2013rza}. In doing
so, however, (i) they ignore the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms could easily dominate over the tidal terms themselves, since there 
are unknown point-particle terms at orders {\it lower} than that at which tidal
effects enter~\cite{Yagi:2014}.
\red{
Make clear the distinction between Fischer matrix studies and Bayesian ones. 
Explain the differences. this is important as it sets the stage for for what is
new in this paper.
} 
Other studies that
use complete numerical simulations or derived models either use Fisher matrix 
to estimate $\lambdans$ measurement error~\cite{Lackey2011,Lackey:2013axa},
which is not reliable at realistic signal-to-noise ratios 
(SNR)~\cite{Vallisneri:2007ev}, or are limited in the binary parameter space
they sample~\cite{Foucart:2013psa}.

In this paper we study the measurability of neutron star's tidal deformability
from GWs emitted by {\it disruptive} NSBH mergers using aLIGO. We also probe how tidal 
effects affect the estimation of other binary parameters for the same class
of systems. This study improves upon previous work in many ways. First, we
include tidal effects during inspiral and merger in a consistent way, by using
the waveform model of Ref.~\cite{Lackey:2013axa} (henceforth ``LEA+'').
\red{
(describe meaning of the acronym LEA+)
}
Second, we include the effect
of black hole spin on tidal GW signals, in addition to the effect of BH mass 
and NS's deformability itself. Third, we perform a fully Bayesian 
analysis, instead of using a Fisher matrix approximation. Fourth, we explore 
how our measurement errors decrease as we gain information from multiple 
(realistic) events.

We first probe the effect of ignoring tidal effects in parameter 
estimation on the recovery of non-tidal binary parameters, such as 
component masses and spins. This is the case of current and planned aLIGO
efforts.
To do so, we use the LEA+ model to generate a set of realistic signals, and
use that as data from which we estimate binary mass and spin parameters with
a Bayesian Monte-Carlo algorithm, using non-tidal (BBH) waveforms as filters.
% 
\red{
The following numerical details disrupt the flow of the text. MOVE after
description of analyses.
}
\prayush{
As in rest of the paper, we restrict here our parameter space to span 
mass-ratios $q:=\mbh/\mns\in[2,5]$, black hole spin (aligned with orbit)
$\chibh\in[-0.5, +0.75]$, and dimensionless neutron star tidal deformability 
$\lambdans:= G\left(\frac{c^2}{G \mns}\right)^5\lambda \in[500, 2000]$.
The mass of NS for signals (and not templates) is fixed to $1.35M_\odot$ 
throughout~\cite{stellarcollapsemass}, and is taken to be 
non-spinning~\cite{Miller:2014aaa}.
Each signal is rescaled to a range of SNR values $\rho\in[10, 70]$.
}
\red{
MOVE the following to SUMMARY. Here, keep 1-2 sentences:
For individual events, ignoring tidal effects will affect mass-estimation
only marginally; only for very loud signals (SNRs $\gtrsim 30$) the systematic
biases exceed the statistical uncertainties. Furthermore, detection searches
can ignore tidal effects without loss of sensitivity.
}
\prayush{
We find that non-inclusion of tidal effects leads to systematic biases in
the measurement of non-tidal parameters. Consider first the binary chirp 
mass $\mchirp:=M\eta^{3/5}$ where $M$ is the binary total mass, and 
$\eta:=\mbh\mns/M^2$ is the dimensionless mass-ratio. We find that for
realistic (moderate) SNRs, i.e. $\rho\lesssim 30$, not including tidal
effects introduces a barely measurable bias that remains below $50\%$
of the statistical uncertainty of the $\mchirp$ measurement itself.
For higher signal strengths, we find that the underlying statistical
uncertainty decreases enough for the systematic bias to become comparable
to it for $\rho=50-70$ and higher. This is expected because chirp mass is fairly
accurately determined by the inspiral, where tidal effects are weak.
Next, we consider the mass-ratio $\eta$. We find that only if the NS is quite
deformable (with $\lambdans=2000$), {\it and} has a companion BH whose mass
is $\mbh\lesssim 4.5M_\odot$ (i.e. in the astrophysical ``mass-gap''~\cite{
Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}),
will the systematic bias in $\eta$ measurement become comparable to its
statistical uncertainties at SNRs $\rho\leq 30$. For SNRs $\rho\gtrsim30$,
however, not including tidal effects is very likely to {\it significantly} bias
the mass-ratio estimate for NSBH binaries. Finally, we consider the spin on the
black hole $\chibh$. We find that $\chibh$ measurements are affected in a similar
way to $\eta$. By not including tidal effects, we compromise our measurement of
$\chibh$ significantly only if signal $\rho\gtrsim 30$. 
%
Overall, we learn an important thing. Bayesian parameter estimation algorithms
aimed at identifying NSBH binaries with low latency in order to alert 
electromagnetic (EM) observers in time to establish a possible coincident 
observation of a GRB~\cite{2012A&A...541A.155A,Singer:2014qca,Singer:2015ema,
Pankow:2015cra} can use BBH templates to look for NSBH systems with aLIGO.
This is so because the primary requirement of the EM follow-up effort is rapid
classification of a binary as an NSBH system, which can be achieved just as 
easily with BBH templates, purely on the basis of the smaller component's mass.
Another conclusion we draw is that detection searches are unaffected by the
choice of ignoring tidal effects in matched-filtering templates.
}

\red{
Somewhat reduce discussion of results. Discuss in order of probability, starting
with low SNR, ending with exceptional events.
}
\prayush{
Next, we study the ability of aLIGO to constrain neutron star tidal 
deformability with a single observation of coalescing NSBHs. For this, we
use the same setup and parameter choices for signal waveforms as before, but
replace the template model with one that includes tidal effects from inspiral
through to merger~\cite{Lackey:2013axa} (i.e. LEA+). First of all, this reduces
the biases in the measurement of non-tidal parameters. Second, as we can see
from Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, at signal-to-noise
ratios $\simeq 20-30$, only under the most favorable circumstance (including a
BH with mass within the ``mass-gap'') can we put a factor of $1.5-2$ bound on
$\lambdans$ at $90\%$ confidence level. For similar signals with 
$\rho\gtrsim 30$ we can constrain $\lambdans$ within $\pm 50\%$ of its true
value. For more realistic SNRs and BH parameters, a single measurement is 
for $\lambdans$ is accompanied with a $\sim\pm150+\%$. uncertainty.
% While these
% results are expected, multiple realistic observations is the next window that 
% we explore in order to obtain more information on $\lambdans$. These results 
% are detailed in Sec.~\ref{s1:PEwithNS}.
}

\red{
BEgin this paragraph by pointing out that EOS is universal among all NSs. 
Therefore, one universal $\lambdans$ vs $\mns$ curve. Therefore, by combining
multiple BHNS events, expect to measure $\lambdans$ better than with any 
single BHNS event.
\prayush{
Multiple realistic observations deliver far more information than what a single
observation furnishes, as has been shown the case to be for binary neutron
stars~\cite{DelPozzo:13}. As our final study, we perform a fully-Bayesian 
analysis of how our estimation of $\lambdans$ changes as we accumulate more and
more detections.
\red{
MOVE into later sections. Keep here only 1-2 informational sentences.
\prayush{
The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
\textbf{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}.
\red{
MOVE to SUMMARY, only keeping 2-4 sentences here. Also, paradigm A/B barely
matters. Therefore, first describe the common results, then point out how the
presence/absence of mass gap modifies them.
\prayush{
In addition, we probe two paradigms, one
that allows for BH masses to lie within the astrophysical mass-gap (paradigm
A), and one that does not (paradigm B).
For paradigm A, we find the following: (i) for the softer equations of state
that result in less deformable neutron stars, $15-20$ detections are sufficient
for our median measured value of $\lambdans$ to lie within $10\%$ of the true
value. (ii) For NSBH populations with more deformable NSs
($\lambdans\geq 1000$), the same is achievable within as few as $10$ realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans\geq 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs do indeed have
masses between $\leq 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
For paradigm B, we find the information accumulation to be somewhat slower.
While the quantitative inferences for $\lambdans\geq1000$ populations are
not affected significantly, we find $\mathcal{O}(10-20)\%$ more events are
required to attain the same measurement accuracy as under paradigm A.
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, with a few years'
time-scale.

Finally, we note that our results are applicable for the design sensitivity
LIGO instruments, that are expected to come online in $2018-19$.
\red{
MOVE towards end of SUMMARY, into ``Future work''.
}
\prayush{
Also note
that more a recent work that improves upon the waveform model we use in this
study has been published~\cite{Pannarale:2015jka}, but it provides only an 
amplitude model which needs to be (in future) augmented with a compatible
phase model. In addition, accuracy of the underlying numerical simulations 
used to calibrate the waveform model used here have not been verified against 
independent codes so far.
It is therefore difficult to assess the combined modeling error and its effect
on our results. they are, therefore, limited by the limitations of our
waveform model. However, we expect the combined effect of modeling errors to
not change our qualitative conclusions.
}
